\chapter*{Acknowledgements}
\label{acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

\newpage

\noindent \textbf{Abstract}

\noindent \textbf{Background and objective}: In najaar 2007, a world-renowned scientist and all around brilliant lady, by the name of Professor Dr. Ellen Kampman, placed an advertisement for a PhD student in the looking-for-a-PhD section on a popular academic work-related website. A visitor to the site, I, Audrey Jung, was successful in securing the PhD position after serious vetting procedures that may or may not have included anthropometric measurements, colonoscopies, biomarker analyses, and completion of multiple dietary and lifestyle questionnaires. The objective for the following years would be for me to complete my thesis (under the daily coaching of Ellen Kampman) with the least number of casualties and collateral damage.

\noindent \textbf{Subjects and methods}: We developed a novel and unique study design that can never be replicated nor reproduced. Our population consisted of a daily supervisor (n=1 (EK)), co-supervisors (n=2 (FN\&HB)), and a PhD student (n=1 (AJ)) with a number of collaborators/coauthors/lab technicians/research assistants (n=lots) and group members (n=lots too).

\noindent Essential to our project were participants from four studies -- the Nijmegen Biomedical Study, the FACIT study, the POLIEP-follow up study, and the GEOLynch study. Without their enthusiastic participation, this thesis would not have been possible. Thank you so very much.

\noindent Funding for this research was kindly provided by the Dutch Cancer Society (KWF).

\noindent The all around brilliant lady, Ellen Kampman, thank you for giving so much of your time to this project. A heartfelt thanks for your unwavering support of me throughout this endeavor. You are so inspiring not only professionally, but also in the elegant ways in which you combine all the different aspects of your life. Thank you for the talks, for all your care, all the dinners, for everything. It has been a privilege working with you.

\noindent Fokko Nagengast and Henk Blom, your mentorship was invaluable. Thank you for being a part of this project and for always helping answer my numerous questions.

\noindent The Department for Health Evidence was the main setting for work done on this thesis, and the kindness of all the people at this department helped to ensure its completion. Thank you very much for all your immense support.

\noindent I also spent quite some time at the Division of Human Nutrition at Wageningen University, especially with the Nutrigenomics group. Thank you to all at the Division for your hospitality.

\noindent A shout out to Bart Kiemeney, who was not only always able to come up with clever suggestions for issues pertaining specifically to this thesis but was also genuinely concerned about my well-being and personal situation. Thank you, Bart, for your care, openness, and your friendliness.

\noindent Astrid van Alst and Simone van Wezel at the Department for Health Evidence made sure that I did not end up homeless and worked tirelessly on sorting out administration matters. Thank you!

\noindent At the Department of Pathology, Patricia van Beuningen and Els Schut taught me how to use the PALGA database and spent countless days and months helping me collect tissues from the different hospitals. Elisa Vink-B{\"o}rger showed me how to cut paraffin-embedded tissues, do IGF staining, and extract DNA from tissues. She also assessed the histopathology of the many tissues. Thank you all!

\noindent Thank you to the other (former and current) members of Ellen's team with whom I have had very fun times working: Akke Botma, Dieuwertje Kok, Renate Heine-Br{\"o}ring, Fr{\"a}nzel van Duijnhoven, and Renate Winkels. Akke, it was such super duper fun times with my partner in crime. I can only dream of becoming a great scientist like you one day! Dieuwertje, thank you for your very helpful comments on my introduction. Renate H, thank you for all the nice chats! Fr{\"a}nzel, the Lynch paper was a blast! I learned so much from you! Renate W, our meetings were always so enjoyable, and you hosted such a tasty dinner at your place.

\noindent Thank you so much to Jan Harryvan for your expertise in the lab, in particular with DNA isolation and genotyping. Thank you Sandra Heil for your instruction on the ins and outs of DNA isolation and subsequent DNA methylation using the LUMA method.

\noindent Much thanks to Karin Borgonjen and Saskia Meijboom at Dietetics for spending so much time making the nutrient databases.

\noindent Wu Ming, thank you for all the helpful talks. You have great wisdom. Serendipity made me your Master's student, and from there, a friendship was forged. Thank you always for making time for your former student. ??!

\noindent Without Ursula Oldenhof, data collection and organisation for these studies would have been unachievable. Thank you so much Ursula for your help with sending, entering and organising FFQs, all the blood collection, and accompanying me to the scary tissue archives of certain hospitals to obtain samples.

\noindent Wim Lemmens, you are a superstar data manager! And a SAS whiz. It has been an honour. To everyone else with whom I worked closely on the NBS: Martin den Heijer, Tessel Galesloot, Sita Vermeulen, your valuable contributions are gratefully appreciated. Much thanks also to Ton de Haan, for letting me bug you with incessant questions about statistics.

\noindent Maria van Vugt, you are Speedy Gonzales with helping to find information from patients. Thank you!

\noindent Thank you to the members of the reading committee Nicoline Hoogerbrugge, Iris Nagtegaal, and Cornelia Ulrich for your time. Neli, thank you so much as well for taking the time out of your busy schedule to supervise me at the Hutch. I learned so much from you, and it was wonderful to be part of such an incredible team.

\noindent My gratitude and appreciation go to Iris van Rooij and Carolien Lute for agreeing to be my paranymphs. Dear Iris, you were my first officemate, and thank you too for extending a hand as my ``Dutch buddy'' and for sharing such nice dinners with me! Good times were had with Carolien Lute (who moonlights as a personal chauffeur) pipetting and pyrosequencing in the labs all the livelong days!

\noindent Thank you to past and glorious officemates: Judith, Parisa, Janine, Gijs, Martine, and Marcia, who with their good cheer, made things so much more amusing!

\noindent Thank you to all the coauthors for your insight, patience, and for making the manuscripts exponentially better: Jane Durga, Jacqueline de Graaf, Suzanne Holewijn, Jan Kleibeuker, Frans Kok, Robert Kok, Gry Kvalheim, {\O}ivind Midttun, Yvo Smulders, Wilma Steegenga, Per Ueland, Hans Vasen, and Petra Verhoef. Also thank you to Paul Haggarty, who was instrumental in making the book chapter happen.

\noindent To new friendships made: Georgeta, Eleftheria, Nina, Sonja and Luc, Arun Kommadath, Mark Roberts, Basav and Shital, Andreea, Hilario, Deniz, the Lost-in-Nijmegeners, and the friendly neighbourhood peeps. You guys are awesome. Thank you to Jen Wen and Angus, QQ and Alex, Miguel, Brad, Brian, and Jordan Sharif for travelling so far for lovely visits. You guys are awesome too.

\noindent Pro Jana Mrkose, kter{\' y} pro mne byl obrovskou oporou p{\v r}i dokon{\v c}ov{\' a}n{\' i} diplomov{\' e} pr{\' a}ce.
Jsi t{\v r}{\' i}da. B{\accent27'u}h ti {\v z}ehnej.

\noindent To my beautiful family: mom, dad, Adrienne, paternal grandmother, paternal grandfather, maternal grandfather, maternal grandmother, all of whom loved and sacrificed so much for me. This thesis is yours and not my own. Thank you for everything. And after all these years apart, we can finally hang out more now!

\noindent \textbf{Results}: This thesis! The thesis is complete with only few casualties and collateral damage, nothing that cannot be fixed with tea and stroopwafels. We observed that the relationship between me carrying out a PhD project and successful completion of the thesis depended heavily on the patience, involvement, and helpfulness of Ellen Kampman, Fokko Nagengast, and Henk Blom. And also on numerous other factors such as beautiful friends, family, colleagues, and collaborators.

\noindent \textbf{Conclusions}: A journey of this magnitude is never taken alone. Thank you to everyone who in your own beautiful way joined in for the ride. To those whom I may have forgotten to mention, please blame it on my head and not my heart.
